---
engine: knitr
webr:
  show-startup-message: false
  show-header-message: false
filters:
  - webr
toc: false
---

## Introduction

This is a web-enabled code cell, powered by [quarto-webr extension](https://github.com/coatless/quarto-webr), where you have the freedom to play around and write any R code you desire.

By default, the quarto-webr extension avoids loading or requesting additional packages. You can add extra packages when opening the document or on a per-code cell basis. To see available packages, execute the following R code (using WebR or just R):

```{r}
#| label: package
#| echo: true
#| eval: false
available.packages(repos="https://repo.r-wasm.org/", type="source")
```

Packages may also be installed inside of a code cell through the built-in webr::install() function. For example, to install ggplot2, you would need to use: 

```{r}
#| label: webr_install
#| echo: true
#| eval: false
webr::install("ggplot2")
```

## Playground
Write your code here. 

```{webr-r}
# Create a sequence of values from -π to π with a step of 0.1
x <- seq(-pi, pi, 0.1)

# Plot the sine function (sin(x)) on the graph
# Overlaying Graphs
plot(x, sin(x),
     main = "Overlaying Graphs",   # Set the main title of the plot
     ylab = "",                    # Set the label for the y-axis (empty in this case)
     type = "l",                   # Draw a line plot
     col = "blue"                  # Set the color of the line to blue
)

# Add the cosine function (cos(x)) to the existing plot
lines(x, cos(x), col = "red")      # Draw a line for the cosine function with a red color

# Add a legend to the plot to distinguish between sin(x) and cos(x)
legend("topleft",                 # Position the legend at the top-left of the plot
       c("sin(x)", "cos(x)"),     # Names of the lines to be displayed in the legend
       fill = c("blue", "red")    # Fill colors for the corresponding lines in the legend
)

```

